---
title: "Data Visualizations - A Practical Introduction by Kieran Healy"
author: "Philipp Kollenda"
date: "2019-01-02"
output: html_document
---
   
## Data Visualization Notes

```{r setup, include=FALSE}

## The include=FALSE option tells R to run, but not display the 
## code in this chunk.
## By defult, show code for all chunks in the knitted document,
## as well as the output. To override for a particular chunk
## use echo = FALSE in its options.
knitr::opts_chunk$set(echo = TRUE) 

## Set the default size of figures
knitr::opts_chunk$set(fig.width=8, fig.height=5)  

## Load the libraries we will be using
library(gapminder)
library(here)
library(socviz)
library(tidyverse)

```

## Make a Plot

```{r, message=FALSE, warning=FALSE}
## General strategy: First define the object for ggplot and assign it a name.
## Then add geom_*'s etc. additively and display immediately.
p <- ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp))

## A first version of a relatively polished plot
p + geom_point(alpha = 0.3) + 
    geom_smooth(method = "lm") + 
    scale_x_log10(labels = scales::dollar) + 
    labs(x = "GDP Per Capita", y = "Life Expectancy in Years", 
         title = "Economic Growth and Life Expectancy",
         subtitle = "Data points are country-years",
         caption = "Source: Gapminder")

```

The **color** aesthetic affects the appearance of lines and points, the **fill** aesthetic
controls the appearance of the filled areas of bars, polygons or, in this case,
the interior of the smoothers's standard error ribbons.
```{r, fig.height=9, fig.width=12, message=FALSE, warning=FALSE}
## Note that within the curly brackets we can set options to be different from the default.
p <- ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp, 
    color = continent, fill = continent))

p_out <- 
  p + geom_point(alpha = 0.3) + 
    geom_smooth(method = "lm") + 
    scale_x_log10(labels = scales::dollar) + 
    labs(x = "GDP Per Capita", y = "Life Expectancy in Years", 
         title = "Economic Growth and Life Expectancy",
         subtitle = "Data points are country-years",
         caption = "Source: Gapminder")

## Save via ggsave and, if the plot is assigned a name, specify plot = ...
## Vector based graphs (.pdf / .svg) are easier to resize, but larger than raster based graphs (.png / .jpg)
ggsave(filename = here("Kieran Healy, Data Visualization Book", "figures", "gdp_lifeexp_chapter3.pdf"), plot = p_out, 
       height = 9, width = 16, units = "in")
```


## Show the Right Numbers

Use `facet_wrap()` to split data by group and create separate mini-plots for each group.

```{r, facets, message = F, warning = F}
p <- ggplot(data = gapminder, aes(x = year, y = gdpPercap))
p + geom_line(color = "gray70", aes(group = country)) +
    geom_smooth(size = 1.1, method = "loess", se = FALSE) + 
    scale_y_log10(labels = scales::dollar) + 
    facet_wrap(facets = vars(continent), ncol = 5) +
    labs(x = "Year", y = "GDP per capita", 
         title = "GDP per capita on Five Continents")
```
Alternatively, you can use `facet_grid()` to split groups into sub-plots with more
control, along a true 2-dimensional grid. To illustrate, we use data from the `gss_sm`
dataset which contains questions from the 2016 General Social Survey.

```{r, gss_data}
p <- ggplot(data = gss_sm, aes(x = age, y = childs))
p + geom_point(alpha = 0.2) +
    geom_smooth() +
    facet_grid(rows = vars(sex), cols = vars(race))

```

### stat_* functions and their relation to geom's
Every `geom_*` function has an asscociated `stat_*` function that it uses by default.
The reverse is also true, every `stat_*` function has an associated `geom_*` function
that it will plot by default. For example, the `geom_bar()` function automatically
calls the `stat_count()` function which calculates the values for the count and the 
proportion. We can then call these values in the mapping through `y = ..prop..`.

```{r, stat}
p <- ggplot(data = gss_sm, mapping = aes(x = bigregion))
## Note that we need to add group = 1 to tell R to ignore the grouping by the x-variable
## when calculating the proportions.
p + geom_bar(aes(y = ..prop.., group = 1))

## If we want to additionally color the bars according to their category we use the
## fill aesthetic (the color aesthetic would only create a colored border).
## Note that the reader needs no legend, because the categories are already encoded
## in the x axis. We can turn of guidance for an aesthetic with the guides() function.
p + geom_bar(aes(fill = bigregion)) +
    guides(fill = FALSE)

## Let's display what the proportion of each religion within each state is.
## Of all people in the Northeast, what proportion are Catholic, etc... We can do 
## this two ways, by using the same bar diagram and positioning the different religions
## side-by-side...
p <- ggplot(data = gss_sm, mapping = aes(x = religion, fill = bigregion))
p + geom_bar(position = "dodge", mapping = aes(y = ..prop.., group = bigregion))

## While this works, it makes it unintuitive to compare bars across different groups.
## It is also not clear immediately which group (religion or region) sum up to 1.
## In this case, it is probably easier to let the bar's divide by religion and facet
## by region. We then do not need the fill aesthetic.

p <- ggplot(data = gss_sm, mapping = aes(x = religion))
p + geom_bar(mapping = aes(y = ..prop.., group = bigregion)) + facet_wrap(vars(bigregion))
```

### The identity stat and position
You can call `stat = "identity` to say that ggplot should perform no calculation
on the data. This is helpful to display data as is without geom_*'s performing their
default calculations of summary statistics.

```{r}

p <- ggplot(data = titanic, aes(x = fate, y = percent, fill = sex))
p + geom_bar(position = "dodge", stat = "identity")

## An alternative is the geom_col() function which is the same as geom_bar with the
## default stat = "identity".
# p + geom_col(position = "dodge")
```
Similarly, one can set the position argument to `position = "identity"` to tell
ggplot to display the data *exactly where it is*. This can be used (with ordered data)
to use bar plots as alternatives to line plots.

```{r}

p <- ggplot(data = oecd_sum, aes(x = year, y = diff, fill = hi_lo))
p + geom_col() +
    guides(fill = FALSE) +
    labs(x = NULL, y = "Difference in Years", 
         title = "The US Life Expectancy Gap",
         subtitle = "Difference between US and OECD average life expectancies, 1960-2015",
         caption = "Data: OECD. After a chart by Christopher Ingraham, Washington Post, December 27th 2017.")

```
### Exploring alternatives to the histogram

```{r}

p <- ggplot(data = gapminder, aes(x = gdpPercap))
p + geom_bin2d(aes(y = lifeExp), bins = c(25, 20)) +
    scale_x_log10(labels = scales::dollar)

```



## Graph Tables, Make Labels, Add Notes

```{r}

```

## Work with Models

```{r}

```

## Draw Maps

```{r}

```


## Refine your Plots

```{r}

```


