---
title: "Data Visualizations - A Practical Introduction by Kieran Healy"
author: "Philipp Kollenda"
date: "2019-01-02"
output: 
  html_document:
    toc: true
---
   
## Data Visualization Notes

```{r setup, include=FALSE}

## The include=FALSE option tells R to run, but not display the 
## code in this chunk.
## By defult, show code for all chunks in the knitted document,
## as well as the output. To override for a particular chunk
## use echo = FALSE in its options.
knitr::opts_chunk$set(echo = TRUE) 

## Set the default size of figures
knitr::opts_chunk$set(fig.width=8, fig.height=5)  

## Load the libraries we will be using
library(gapminder)
library(here)
library(socviz)
library(tidyverse)

```

## Make a Plot

```{r, message=FALSE, warning=FALSE}
## General strategy: First define the object for ggplot and assign it a name.
## Then add geom_*'s etc. additively and display immediately.
p <- ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp))

## A first version of a relatively polished plot
p + geom_point(alpha = 0.3) + 
    geom_smooth(method = "lm") + 
    scale_x_log10(labels = scales::dollar) + 
    labs(x = "GDP Per Capita", y = "Life Expectancy in Years", 
         title = "Economic Growth and Life Expectancy",
         subtitle = "Data points are country-years",
         caption = "Source: Gapminder")

```

The **color** aesthetic affects the appearance of lines and points, the **fill** aesthetic
controls the appearance of the filled areas of bars, polygons or, in this case,
the interior of the smoothers's standard error ribbons.
```{r, fig.height=9, fig.width=12, message=FALSE, warning=FALSE}
## Note that within the curly brackets we can set options to be different from the default.
p <- ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp, 
    color = continent, fill = continent))

p_out <- 
  p + geom_point(alpha = 0.3) + 
    geom_smooth(method = "lm") + 
    scale_x_log10(labels = scales::dollar) + 
    labs(x = "GDP Per Capita", y = "Life Expectancy in Years", 
         title = "Economic Growth and Life Expectancy",
         subtitle = "Data points are country-years",
         caption = "Source: Gapminder")

## Save via ggsave and, if the plot is assigned a name, specify plot = ...
## Vector based graphs (.pdf / .svg) are easier to resize, but larger than raster based graphs (.png / .jpg)
ggsave(filename = here("Kieran Healy, Data Visualization Book", "figures", "gdp_lifeexp_chapter3.pdf"), plot = p_out, 
       height = 9, width = 16, units = "in")
```


## Show the Right Numbers

Use `facet_wrap()` to split data by group and create separate mini-plots for each group.

```{r, facets, message = F, warning = F}
p <- ggplot(data = gapminder, aes(x = year, y = gdpPercap))
p + geom_line(color = "gray70", aes(group = country)) +
    geom_smooth(size = 1.1, method = "loess", se = FALSE) + 
    scale_y_log10(labels = scales::dollar) + 
    facet_wrap(facets = vars(continent), ncol = 5) +
    labs(x = "Year", y = "GDP per capita", 
         title = "GDP per capita on Five Continents")
```
Alternatively, you can use `facet_grid()` to split groups into sub-plots with more
control, along a true 2-dimensional grid. To illustrate, we use data from the `gss_sm`
dataset which contains questions from the 2016 General Social Survey.

```{r, gss_data}
p <- ggplot(data = gss_sm, aes(x = age, y = childs))
p + geom_point(alpha = 0.2) +
    geom_smooth() +
    facet_grid(rows = vars(sex), cols = vars(race))

```

### stat_* functions and their relation to geom's
Every `geom_*` function has an asscociated `stat_*` function that it uses by default.
The reverse is also true, every `stat_*` function has an associated `geom_*` function
that it will plot by default. For example, the `geom_bar()` function automatically
calls the `stat_count()` function which calculates the values for the count and the 
proportion. We can then call these values in the mapping through `y = ..prop..`.

```{r, stat}
p <- ggplot(data = gss_sm, mapping = aes(x = bigregion))
## Note that we need to add group = 1 to tell R to ignore the grouping by the x-variable
## when calculating the proportions.
p + geom_bar(aes(y = ..prop.., group = 1))

## If we want to additionally color the bars according to their category we use the
## fill aesthetic (the color aesthetic would only create a colored border).
## Note that the reader needs no legend, because the categories are already encoded
## in the x axis. We can turn of guidance for an aesthetic with the guides() function.
p + geom_bar(aes(fill = bigregion)) +
    guides(fill = FALSE)

## Let's display what the proportion of each religion within each state is.
## Of all people in the Northeast, what proportion are Catholic, etc... We can do 
## this two ways, by using the same bar diagram and positioning the different religions
## side-by-side...
p <- ggplot(data = gss_sm, mapping = aes(x = religion, fill = bigregion))
p + geom_bar(position = "dodge", mapping = aes(y = ..prop.., group = bigregion))

## While this works, it makes it unintuitive to compare bars across different groups.
## It is also not clear immediately which group (religion or region) sum up to 1.
## In this case, it is probably easier to let the bar's divide by religion and facet
## by region. We then do not need the fill aesthetic.

p <- ggplot(data = gss_sm, mapping = aes(x = religion))
p + geom_bar(mapping = aes(y = ..prop.., group = bigregion)) + facet_wrap(vars(bigregion))
```

### The identity stat and position
You can call `stat = "identity` to say that ggplot should perform no calculation
on the data. This is helpful to display data as is without geom_*'s performing their
default calculations of summary statistics.

```{r}

p <- ggplot(data = titanic, aes(x = fate, y = percent, fill = sex))
p + geom_bar(position = "dodge", stat = "identity")

## An alternative is the geom_col() function which is the same as geom_bar with the
## default stat = "identity".
# p + geom_col(position = "dodge")
```
Similarly, one can set the position argument to `position = "identity"` to tell
ggplot to display the data *exactly where it is*. This can be used (with ordered data)
to use bar plots as alternatives to line plots.

```{r}

p <- ggplot(data = oecd_sum, aes(x = year, y = diff, fill = hi_lo))
p + geom_col() +
    guides(fill = FALSE) +
    labs(x = NULL, y = "Difference in Years", 
         title = "The US Life Expectancy Gap",
         subtitle = "Difference between US and OECD average life expectancies, 1960-2015",
         caption = "Data: OECD. After a chart by Christopher Ingraham, Washington Post, December 27th 2017.")

```
### Exploring alternatives to the histogram

```{r}

p <- ggplot(data = gapminder, aes(x = gdpPercap))
p + geom_bin2d(aes(y = lifeExp), bins = c(25, 20)) +
    scale_x_log10(labels = scales::dollar)

```



## Graph Tables, Make Labels, Add Notes

```{r}

rel_by_region <- gss_sm %>% group_by(bigregion, religion) %>% 
  summarise(N = n()) %>% mutate(freq = N / sum(N), pct = round((freq*100), 0))

## As a rule, dodged charts can be more cleanly expressed as faceted plots and 
## to prevent long and overlapping labels, coord_flip() is the way to go.
p <- ggplot(data = rel_by_region, mapping = aes(x = religion, y = pct, fill = religion))
p + geom_col(position = "dodge2") +
    labs(x = NULL, y = "Percent", fill = "Religion") +
    guides(fill = FALSE) +
    coord_flip() +
    facet_grid(cols = vars(bigregion))

```

Boxplots are a good way of displaying variation within groups, separately for different
groups. Just like `stat_bar()`, `geom_boxplot()` and its connected `stat_boxplot()` 
calculate summary statistics to display the median, quantiles and outliers.

```{r}

## By default, geom_boxplot orders values on the x-Axis alphabetically, but this is
## usually not the most sensible order. Better to order good on top to bad on bottom.
## You can reorder() by mean (default), median, sd. Here median (consistent with black line in
## boxplot) or mean (consistent with center of mass of box of boxplot) are good.
p <- ggplot(data = organdata, mapping = aes(x = reorder(country, donors, na.rm = T), 
                                            y = donors, fill = world))
p + geom_boxplot() +
    labs(x = NULL) +
    coord_flip() +
    theme(legend.position = "top")

## If there are relatively few observations per group (here country), it is desirable
## to (also) show the raw data with geom_point(). Use jitter to perturb points and 
## avoid overplotting.
p + geom_jitter(aes(fill = NULL, color = world), position = position_jitter(width = 0.15)) +
    labs(x = NULL) +
    coord_flip() +
    theme(legend.position = "top")

```

### Cleveland dotplots. The better bar- or boxplot.
If you are displaying only one point per category (or a summary statistic) it is
often more appearling to use a Cleveland dotplot over a bar plot or a dot-and-whisker
plot over a boxplot. To do this, you first transform the dataset and create the mean
and standard deviation manually and then only plot the mean (as dots) and the standard
deviation (as whiskers).

```{r}

## Note that summarise changed compared to the version in the book. Now scoping
## over variables is done via multiple across() calls.
## The .names below is the default, but would allow for more flexibility. Functions
## are passed as named list.
by_country <- organdata %>% group_by(consent_law, country) %>% 
  summarise(across(is.numeric, list(mean = mean, sd = sd), na.rm = T, .names = "{col}_{fn}"), 
            across(world, first)) %>% 
  ungroup()

p <- ggplot(by_country, aes(x = donors_mean, y = reorder(country, donors_mean), color = consent_law))
p + geom_point(size = 3) +
    labs(x = "Donor procurement rate", y = NULL, color = "Consent Law") + 
    theme(legend.position = "top")

# Add whiskers to show the standard deviation with the geom_pointrange geom. Since
p + geom_point(size = 3) + 
    geom_pointrange(aes(xmin = donors_mean - donors_sd, xmax = donors_mean + donors_sd)) +
    labs(x = "Donor procurement rate", y = NULL, color = "Consent Law") +
    theme(legend.position = "top")

## Instead of color coding the type of consent law, it might be better to facet the plot.
## In that case, keep the y axis free floating to not duplicate country rows.
p <- ggplot(by_country, aes(x = donors_mean, y = reorder(country, donors_mean)))
p + geom_point(size = 3) +
    facet_wrap(vars(consent_law), scale = "free_y", ncol = 1) +
    labs(x = "Donor procurement rate", y = NULL) 

```

### Adding text to plots.

```{r, fig.width = 15, fig.height = 15}

## The ggrepel package and its geom_text_repel() function performs better than the
## geom_text() function and does not overlay labels. 

library(ggrepel)

p <- ggplot(elections_historic, aes(x = popular_pct, y = ec_pct, label = winner_label))
p + geom_vline(xintercept = 0.5, size = 1.4, color = "grey80") +
    geom_hline(yintercept = 0.5, size = 1.4, color = "grey80") +
    geom_point() +
    geom_text_repel() +
    scale_x_continuous(labels = scales::percent) +
    scale_y_continuous(labels = scales::percent) +
    labs(title = "Presidential elections: Popular & electoral college margins",
         subtitle = "1824-2016",
         caption = "Data for 2016 is provisional.",
         x = "Winner's share of popular vote", 
         y = "Winner's share of electoral college votes",
         label = NULL) 
```

```{r}

## Usually, we would want to display labels only for certain points.

p <- ggplot(data = by_country, aes(x = gdp_mean, y = health_mean))
p + geom_point() +
    geom_text_repel(data = subset(by_country, gdp_mean > 25000), 
                    aes(label = country))

organdata$ind <- organdata$ccode %in% c("Ita", "Spa") & organdata$year > 1998
p <- ggplot(organdata, aes(x = roads, y = donors))
p + geom_point(aes(color = ind)) +
    geom_text_repel(data = subset(organdata, ind), mapping = aes(label = ccode)) +
    guides(color = FALSE, label = FALSE)

## It looks nice to fade unlabelled points with alpha. Can be done in aesthetic, but
## maybe more robust to do it in geom_point(). geom_text_repel can create labels 
## based on a combination of values from various columns which is cool!
p + geom_point(data = subset(organdata, ind), alpha = 1) +
    geom_point(data = subset(organdata, !ind), alpha = 0.2) +
    geom_text_repel(data = subset(organdata, ind), mapping = aes(label = paste(ccode, lubridate::year(year), sep = " "))) +
    guides(label = FALSE) +
    labs(caption = "Spain and Italy after 1998 are highlighted, other countries are faded out.")

```
In the previous examples we added labels to the graph, were the labels had to be 
(a combination of) existing values in the dataset, like country-year pairs. Instead
we can add commentary to a plot by annotating it manually. `annotate()` simply tells
R to add **something** to the plot. We specify what should be added with the `geom = ...`
option. 

```{r}

p <- ggplot(organdata, aes(x = roads, y = donors))
p + geom_point() + 
    annotate(geom = "text", label = "A surprisingly high \n recovery rate.",
             x = 157, y = 33, hjust = 0) +
    annotate(geom = "rect", fill = "red", alpha = 0.2,
             xmin = 125, xmax = 155, ymin = 30, ymax = 35) + 
    annotate(geom = "curve", x = 175, y = 31.5, xend = 154, yend = 30, curvature = -0.5, color = "red", alpha = 0.7, arrow = arrow(length = unit(0.5, "cm")))


```
### Exercises

```{r}

elections_historic <- elections_historic %>% filter(win_party %in% c("Dem.", "Rep.")) %>% 
  mutate(ind = year > 1980)
party_colors <- c("#2E74C0", "#CB454A")

p <- ggplot(data = elections_historic, aes(x = popular_pct, y = ec_pct, color = win_party))
p + geom_vline(xintercept = 0.5, size = 1.4, color = "grey80") +
    geom_hline(yintercept = 0.5, size = 1.4, color = "grey80") +
    geom_point(data = subset(elections_historic, !ind), alpha = 0.4) +
    geom_point(data = subset(elections_historic, ind)) +
    geom_text_repel(data = subset(elections_historic, ind), aes(label = winner_label)) +
    scale_x_continuous(labels = scales::percent) +
    scale_y_continuous(labels = scales::percent) +
    scale_color_manual(labels = c("Democrat", "Republican"), values = party_colors) +
    guides(label = F) +
    labs(title = "Presidential elections: Popular & electoral college margins",
         subtitle = "1824-2016",
         caption = "Data for 2016 is provisional.",
         x = "Winner's share of popular vote", 
         y = "Winner's share of electoral college votes",
         color = "",
         label = NULL) +
    theme(legend.position = "top")

```

```{r}

party_colors <- c("#2E74C0", "#CB454A")

p <- ggplot(subset(elections_historic, year > 1940), aes(x = popular_pct, y = reorder(winner_label, popular_pct), color = win_party))
p + geom_vline(xintercept = 0.5) + 
    geom_point() +
    guides(color = F) +
    scale_x_continuous(labels = scales::percent) +    
    scale_color_manual(values = party_colors) +
    labs(x = "Winner's share popular vote", y = NULL) +    
    guides(color = FALSE) +
    annotate(geom = "curve", xend = 0.508, yend = 8.75, x = 0.53, y = 7, 
             arrow = arrow(length = unit(0.5, "cm")), curvature = -0.3) +
    annotate(geom = "text", label = "In 2004, George Bush won Florida \n only after a controversial decision \n of the Supreme Court.",
             x = 0.53, y = 7, hjust = 0)


```


## Work with Models

Get to know the `broom` Package. `tidy()` extracts estimates, standard errors and
confidence intervals. We can change the labels by editing the $term argument.

```{r}

out <- lm(lifeExp ~ gdpPercap + pop + continent, data = gapminder)
out_conf <- tidy(out, conf.int = T) %>% round_df
out_conf <- subset(out_conf, !term %in% c("(Intercept)"))
out_conf$nicelabs <- str_replace(out_conf$term, "continent", "")

p <- ggplot(out_conf, aes(x = reorder(nicelabs, estimate), y = estimate,
                          ymin = conf.low, ymax = conf.high))
p + geom_pointrange() +
    coord_flip() +
    labs(x = "", y = "OLS Estimate")
```
The broom package is powerful in combination with `group_by()` and `nest()` to 
create model fits separately for different groups.

```{r}

fit_ols <- function(df) {
  lm(lifeExp ~ log(gdpPercap), data = df)
}

out_le <- gapminder %>% 
  group_by(continent, year) %>% 
  nest() %>% 
  mutate(model = map(data, fit_ols),
         tidied = map(model, tidy, conf.int = T)) %>% 
  unnest(tidied, .drop = T) %>% 
  filter(term != "(Intercept)" & continent != "Oceania")

p <- ggplot(out_le, aes(x = year, group = continent, color = continent,
                        y = estimate, ymin = conf.low, ymax = conf.high))
p + geom_pointrange(position = position_dodge(width = 1)) +
    scale_x_continuous(breaks = unique(gapminder$year)) +
    labs(x = "", y = "OLS Estimate, log(GDP per Capita)", color = "Continent") +
    theme(legend.position = "top")
    

```

Not all regression estimates translate as straightforward to marginal effects as
regular OLS regressions. For example, for logistic regression (`glm(formula, family = "binomial", ...)`)
we can use the `margins()` package.

```{r}

library(margins)

## Use relevel to define a factor level as the default.
gss_sm$polviews_m <- relevel(gss_sm$polviews, ref = "Moderate")

## Use glm(family = "binomial") for logistic. sex*race creates interaction effects.
out_bo <- glm(obama ~ polviews_m + sex*race, family = "binomial", data = gss_sm)

bo_gg <- as_tibble(summary(margins(out_bo)))
bo_gg$factor <- str_remove(bo_gg$factor, "polviews_m|sex")
bo_gg$factor <- str_replace(bo_gg$factor, "race", "Race: ")

p <- ggplot(data = bo_gg, aes(x = reorder(factor, AME),
                              y = AME, ymin = lower, ymax = upper))
p + geom_pointrange() +
    geom_hline(yintercept = 0, color = "gray80") +
    labs(x = "", y = "Average Marginal Effect") +
    coord_flip()

## It is also nice to plot coniditonal marginal effect:
pv_cp <- cplot(out_bo, x = "sex", draw = F)

p <- ggplot(data = pv_cp, aes(x = reorder(xvals, yvals), y = yvals, ymin = lower, ymax = upper))
p + geom_pointrange() +
    geom_hline(yintercept = 0, color = "gray80") +
    labs(x = "", y = "Conditional effect") +
    coord_flip()

```

Let us now look at survey data explicitly.
```{r}

library(survey)
library(srvyr)
options(survey.lonely.psu = "adjust")
options(na.action = "na.pass")

gss_wt <- subset(gss_lon, year > 1974) %>% 
  mutate(stratvar = interaction(year, vstrat)) %>% 
  as_survey_design(ids = vpsu,
                   strata = stratvar,
                   weights = wtssall,
                   nest = TRUE)

out_mrp <- gss_wt %>% 
  filter(year %in% seq(1976, 2016, by = 4)) %>% 
  mutate(racedeg = interaction(race, degree)) %>% 
  group_by(year, racedeg) %>% 
  summarize(prop = survey_mean(na.rm = T)) %>% 
  separate(racedeg, into = c("race", "degree"), sep = "\\.")

out_grp <- gss_wt %>% 
  filter(year %in% seq(1976, 2016, by = 4)) %>% 
  group_by(year, race, degree) %>% 
  summarize(prop = survey_mean(na.rm = T))

p <- ggplot(data = subset(out_grp, race != "Other"), 
            mapping = aes(x = degree, color = race, fill = race, group = race, 
                          y = prop, ymin = prop-2*prop_se, ymax = prop+2*prop_se))
p + geom_col(position = position_dodge(width = 0.9), alpha = 0.2) +
    geom_errorbar(position = position_dodge(width = 0.9), width = 0.2) +
    scale_x_discrete(labels = scales::wrap_format(10)) +
    scale_y_continuous(labels = scales::percent) +
    scale_color_brewer(type = "qual", palette = "Dark2") +
    scale_fill_brewer(type = "qual", palette = "Dark2") +
    labs(title = "Educational Attainment by Race", 
         subtitle = "GSS 1976-2016",
         fill = "Race", color = "Race",
         x = "", y = "") +
    facet_wrap(vars(year), ncol = 2) +
    theme(legend.position = "top")

## Alternatively, we can use a line plot to better show time trends.
p <- ggplot(subset(out_grp, race != "Other"), aes(x = year, color = race, fill = race, group = race,
                         y = prop, ymin = prop-2*prop_se, ymax = prop+2*prop_se))
p + geom_ribbon(alpha = 0.3, aes(color = NULL)) +
    geom_line() +
    facet_wrap(vars(degree), ncol = 1) +
    scale_y_continuous(labels = scales::percent) +
    scale_color_brewer(type = "qual", palette = "Dark2") +
    scale_fill_brewer(type = "qual", palette = "Dark2") +
    labs(title = "Educational Attainment by Race", 
         subtitle = "GSS 1976-2016",
         fill = "Race", color = "Race",
         x = NULL, y = "") +
    theme(legend.position = "top")

```



## Draw Maps

```{r}

```


## Refine your Plots

```{r}

```


